---
title: "ESM244 Assignment 2 Task 2: Binary Logistic Regression"
author: "Grace Brofman"
date: "2/1/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
In a separate .Rmd saved as a2_task2_firstname_lastname.Rmd, using the Florida palmetto data (palmetto.csv), use binary logistic regression to test feasibility of using variables plant height (height), canopy length (length), canopy width (width), and number of green leaves (green_lvs) to classify whether a palmetto is species Serenoa repens or Sabal etonia. Use code folding and hide all messages & warnings in your knitted HTML. 

Data source: Abrahamson, W.G. 2019. Survival, growth and biomass estimates of two dominant palmetto species of south-central Florida from 1981 - 2017, ongoing at 5-year intervals ver 1. Environmental Data Initiative. https://doi.org/10.6073/pasta/f2f96ec76fbbd4b9db431c79a770c4d5

More information and metadata: https://portal.edirepository.org/nis/metadataviewer?packageid=edi.317.1

```{r}
library(tidyverse)
library(ggfortify)
library(viridis)
library(RColorBrewer)
library(hrbrthemes)
library(patchwork)
library(here)
library(tidyverse)
library(GGally) # for pairs plots
library(broom) # for tidy model outputs in dataframes
library(jtools) # to make quick graphs of model outputs

palmetto <- read_csv(here("data", "palmetto.csv"))
```

### Section 1: Explore the differences in height, canopy length, canopy width, and green leaves for the two species. Add a figure caption to each. Below each data visualization, add a short (~1 sentence) statement with a takeaway from the visualization.

```{r}
# according to metadata, 1 (in species column) refers to "Serenoa repens" and 2 refers to "Sabal etonia"
# see metadata for all other column/variable specifications

palmetto_explore <- palmetto %>% 
  select(species, height, length, width, green_lvs) %>% 
  mutate(species = as.factor(species))  # changing species classification from character to factor
  
palmetto_explore %>% 
  ggpairs(aes(color = species))

# pairs plots show lots of similarity in height and width, and the most visible difference in green leaves followed by length

# visualizing width and height by species
ggplot(data = palmetto_explore,
       aes(x = width,
           y = height)) +
  geom_point(aes(color = species))
# appears to be lots of overlap between the species for height/width

# visualizing height and green leaves by species
ggplot(data = palmetto_explore,
       aes(x = height,
           y = green_lvs)) +
  geom_point(aes(color = species))
# generally, nubmer of green leaves increases with height. species 1 appears to have a greater number of green leaves than species 2

# visualizing canopy length and green leaves by species
ggplot(data = palmetto_explore,
       aes(x = length,
           y = green_lvs)) +
  geom_point(aes(color = species))
# generally, green leaves and canopy length appear to be positively correlated. species 1 tends to have more green leaves (and a broader range of # of leaves)
# 6215 of species 1, 6245 of species 2
```

```{r}
# Palmetto height by species density plot

graph_1 <- ggplot(data = palmetto_explore,
       aes(x = height, group = species, fill = species)) +
  geom_density(alpha = 0.5) +
  theme_ipsum() + # custom theme
  labs(x = "Height (cm)",
       y = "Density",
       title = "Palmetto height density distribution by species") +
  theme(legend.position = c(0.9, 0.6), # moves legend
        legend.direction = "vertical") + # legen orientation
  scale_fill_manual(values = c("#21908c80", "#fde72580"), # custom colors
                    name = "Species",
                    labels = c("Serenoa repens", "Sabal etonia"))
  
```


```{r}
# Palmetto width by species density plot

graph_2 <- ggplot(data = palmetto_explore,
       aes(x = width, group = species, fill = species)) +
  geom_density(alpha = 0.5) +
  theme_ipsum() + # custom theme
  labs(x = "Width (cm)",
       y = "Density",
       title = "Palmetto canopy width density distribution by species") +
  theme(legend.position = c(0.9, 0.6), # moves legend
        legend.direction = "vertical") + # legen orientation
  scale_fill_manual(values = c("#21908c80", "#fde72580"), # custom colors
                    name = "Species",
                    labels = c("Serenoa repens", "Sabal etonia"))
```


```{r}
# Palmetto canopy length by species density plot (length = widest length of the canopy)

graph_3 <- ggplot(data = palmetto_explore,
       aes(x = length, group = species, fill = species)) +
  geom_density(alpha = 0.5) +
  theme_ipsum() + # custom theme
  labs(x = "Canopy length (cm)",
       y = "Density",
       title = "Palmetto canopy length density distribution by species") +
  theme(legend.position = c(0.9, 0.6), # moves legend
        legend.direction = "vertical") + # legend orientation
  scale_fill_manual(values = c("#21908c80", "#fde72580"), # custom colors
                    name = "Species",
                    labels = c("Serenoa repens", "Sabal etonia"))
```



```{r}
# Palmetto leaf number by species density plot

graph_4 <- ggplot(data = palmetto_explore,
       aes(x = green_lvs, group = species, fill = species)) +
  geom_histogram(alpha = 0.5, bins = 19, color = "black") +
  theme_ipsum() + # custom theme
  labs(x = "Green leaf count",
       y = "Density",
       title = "Palmetto green leaf count density distribution by species") +
  theme(legend.position = c(0.9, 0.6), # moves legend
        legend.direction = "vertical") + # legen orientation
  scale_fill_manual(values = c("#21908c80", "#fde72580"), # custom colors
                    name = "Species",
                    labels = c("Serenoa repens", "Sabal etonia"))
```


```{r}
# combining graphs into one using patchwork
(graph_1 + graph_2) / (graph_3 + graph_4)
```

### Section 2: perform binary logistic regression using plant height, canopy length, canopy width and green leaves as predictor variables to understand how they relate to probability of a plant being Serenoa repens or Sabal etonia (make sure you understand which species is the first ‘0’ factor level, and which is ‘1’ - you may want to convert to a factor first, then use the levels() function to check). The output should be a finalized table containing the binary logistic regression model results (at least coefficients, standard errors for the coefficients, and information for significance - consider using broom::tidy() to get you most of the way). 

Binary logistic regression
- plant height
- canopy length
- canopy width
- green leaves
As predictor values for species
```{r}
# species has already been changed to a factor in palmetto_explore, so need to determine which factor is "0" level using levels(palmetto_explore$species)
# no need to use fact_drop() (see lab 3 part 1) because we're already working with just the species we need
# Species "1" (Serenoa repens) is 0 order, Species "2" (Sabal etonia) is 1 order

# setting up generalized linear model
palmetto_blr <- glm(species ~ height + length + width + green_lvs,
                    data = palmetto_explore,
                    family = "binomial")

palmetto_blr # shows intercept/coefficient values for each predictor variable

summary(palmetto_blr) # gives more information - all p-values are significantly < 0.05

# because species 1 is the reference species, the left-hand side of our log linear expression (equation) will the the probability of species 2 (the log odds of the non-zero level)
# we expect on average that as height increases, the odds of it being species 2 goes down (based on the sign), although it's the smallest coefficient of the 4.
# ... as length increases the odds of it being species 2 increase
# ... as width increases, the odds of it being species 2 decrese
# ... as number of green leaves increases, the odds of it being species 2 decrease. This is the largest coefficient of the 4, suggesting it may be the strongest predictor variable (this is also reflected in the graphs)

```

```{r}
# data viz - does it make sense that as number of leaves increases, the odds of being species 2 decreases?
ggplot(data = palmetto_explore,
       aes(x = species,
           y = length)) +
  geom_jitter(aes(color = species))
```
```{r}
# look at actual probabilities using broom::augment()
# What are the actual probabilities of being species 2 for each of the existing observations in our palmettoexplore dataframe?

palmetto_fitted <- palmetto_blr %>% 
  broom::augment(type.predict = "response")

# now added on to df we see more info for our model outcomes
# .fitted column is based on our blr, the probability based on 4 predictor variables, that our palmetto is species 2. 

# plotting fitted probabilities
ggplot(data = palmetto_fitted,
       aes(x = green_lvs,
           y = .fitted)) +
  geom_point(aes(color = species)) +
  geom_smooth(aes(color = species), se = FALSE) +
  labs(x = "Number of green leaves",
       y = "Probability of outcome Species 2")

# we see that there are plenty of species 1 that would be classified as species 2

# use broom::tidy() to convert to table

palmetto_table <- palmetto_blr %>% 
  broom::tidy() %>% 
  rename(coeficients = term) %>% 
  rename ("standard error" = std.error) %>% 
  rename("p-value" = p.value)

palmetto_table

```

### Section 3: A section that evaluates how successfully this model would “classify” a plant as the correct species, using a 50% cutoff (e.g. if the probability is >=50% that it is species A, then it would be classified as species A). You should use broom::augment() to find the probabilities for each plant in the original dataset, then add a column for which species the model (using that 50% cutoff) would classify that plant as using the 4 included predictor variables. The outcome should be a finalized table showing, for each species, how many plants in the original dataset would be correctly classified and how many were incorrectly classified by the model, as well as an additional column with “% correctly classified”. Add a table caption above the table

```{r}
palm_fit <- palmetto_fitted %>% 
mutate(species_name = case_when(species == 1 ~ "Serenoa_repens",
                             species == 2 ~ "Sabal_etonia"))

#palmetto_predict <- palmetto_fitted %>% 
 # broom::augment()


# palm_predict <- palm_fit %>% 
#   mutate(pred_spec = case_when(.fitted >= .5 ~ "Serenoa_repens",
#                                .fitted < .5 ~ "Sabal_etonia")) %>% 
#   mutate(true_pred = case_when(species_name == pred_spec ~ "Correct",
#                                species_name != pred_spec ~ "Incorrect"))
# 
# counting <- palm_predict %>% 
#   group_by(species_name) %>% 
#   count(true_pred)
# 
# count_table <- counting %>% 
#   pivot_wider(names_from = true_pred, values_from = n) %>% 
#   mutate(pct_correct = (Correct / (Correct + Incorrect))*100)

# trying again with switched species names
palm_predict_2 <- palm_fit %>% 
    mutate(pred_spec = case_when(.fitted >= .5 ~ "Sabal_etonia",
                               .fitted < .5 ~ "Serenoa_repens")) %>% 
  mutate(true_pred = case_when(species_name == pred_spec ~ "Correct",
                               species_name != pred_spec ~ "Incorrect"))

counting_2 <- palm_predict_2 %>% 
  group_by(species_name) %>% 
  count(true_pred)

count_table_2 <- counting_2 %>% 
  pivot_wider(names_from = true_pred, values_from = n) %>% 
  mutate(pct_correct = (Correct / (Correct + Incorrect))*100)

# make a kable table
library(kableExtra)

prediction_table <- count_table_2 %>% 
  kable(col.names = c("Species",
                      "Correctly Classified",
                      "Incorrectly Classified", 
                     "Percent Correctly Classified")) %>% 
  kable_styling(bootstrap_options = "striped", 
                full_width = F,
                position = "left") %>% 
  add_header_above(c("Test title" = 4))

prediction_table

```



Data source: Abrahamson, W.G. 2019. Survival, growth and biomass estimates of two dominant palmetto species of south-central Florida from 1981 - 2017, ongoing at 5-year intervals ver 1. Environmental Data Initiative. https://doi.org/10.6073/pasta/f2f96ec76fbbd4b9db431c79a770c4d5






